Errorless irrationality: removing error-driven components from the inverse base-rate effect paradigm

### 2023-01-10 docs/manuscript.tex
continue working on discussion

\cogscifinalcopy % Uncomment this line for the final submission

@@

In this study, we tested a central assumption of the most prominent theories of the IBRE.
This central assumption was prediction error.
In our first attempt, we implemented an observational learning version of the canonical IBRE procedure.
This meant that features and category labels appeared on the screen at the same time.
Participants learned about categories by reading complete sentences that describe what symptoms people exhibit and what diseases they have.
The experiment included no feedback and required no responses from participants during training.
From a theoretical perspective, there was no opportunity for making an explicit error, therefore the condition for prediction error.
Regardless, we observed the inverse base-rate effect.
One limitation of this approach was that there are assumed causal relationships between features (symptoms) and labels (diseases).
These relationships can predispose participants to make feature-to-label predictions, which will results in prediction error and attentional relocation.
In our second attempt, we further removed the causal relationship between features and labels by changing the stimuli and their presentation.
Here, participants saw nothing but an arrangement of geometric shapes, where previous category labels were treated as features.
There were no causal links between features and labels.
When participants needed to complete incomplete arrangements of these shapes, they still exhibited a rare bias on $BC$ trials.
So, we still observed the IBRE.

% LD: so it seems that johansen et al treated asymmetric representation as an experimental condition and not as a theoretical construct
% LD: I think this is a problem, because asymmetric representation is hypothesized to be a asymmetry in the way the two categories represented
- asymmetric representation as an assumption

@@

We made available the two experiments written in javascript, the analysis code, the raw data, and all other supplementary materials.
Repositories related to Experiment 1 can be found on \href{https://osf.io/auwvt/?view_only=2dc8384074fa4bcf9f2e3937fdaee2b4}{OSF}, and \href{https://github.com/lenarddome/ply216-observational-ibre}{GitHub}.
Repositories related to Experiment 2 can be found on \href{https://osf.io/2tmc4/?view_only=489ebc888ef84a6b9b904072cfbe74df}{OSF} and \href{https://github.com/lenarddome/ply222-non-causal-ibre}{Github}.
The main repository that includes this manuscript and links to the materials for the two experiments can be found on \href{https://github.com/lenarddome/pu093-errorless-attentionless-ibre}{GitHub}.

>>>>

% \cogscifinalcopy % Uncomment this line for the final submission

@@

In this study, we tested a central assumption of the most prominent theories of the IBRE.
This central assumption was prediction error.
In our first attempt, we implemented an observational learning version of the canonical IBRE procedure.
This meant that features and category labels appeared on the screen at the same time.
Participants learned about categories by reading complete sentences that describe what symptoms people exhibit and what diseases they have.
The experiment included no feedback and required no responses from participants during training.
From a theoretical perspective, there was no opportunity for making an explicit error, therefore the condition for prediction error.
Regardless, we observed the inverse base-rate effect.
One limitation of this approach was that there are assumed causal relationships between features (symptoms) and labels (diseases).
These relationships can predispose participants to make feature-to-label predictions, which will results in prediction error and attentional relocation.
In our second attempt, we further removed the causal relationship between features and labels by changing the stimuli and their presentation.
Here, participants saw nothing but an arrangement of geometric shapes, where previous category labels were treated as features.
There were no causal links between features and labels.
When participants needed to complete incomplete arrangements of these shapes, they still exhibited a rare bias on $BC$ trials.
We still observed the IBRE.

The two experiments suggest that the necessary conditions to observe the IBRE are fewer than previously established.
In Experiment 2, the only remaining conditions are the two uniquely predictive features, an overlapping feature, sequential presentation and the base rate.
\cite{johansen2007paradoxical} gave evidence that even the overlapping features are not necessary.
In their design, common category exemplars were made up of two features, while rare exemplars only had a single feature.
This can mean that common exemplars need to be composed of at least two features, while the rare exemplar must have only one feature.
If the rare exemplar has one uniquely predictive feature and another feature that is not shared with the other category, we do not observe the effect.
This is the shared-cue effect \cite{kruschke2001inverse, wills2014attention}.
One can hypothesise that a shared cue is only a sufficient condition.
In addition, one can hypothesise that in the absence of a shared feature, common exemplars must contain more features than rare exemplars.

% LD: so it seems that Johansen et al treated asymmetric representation as an experimental condition and not as a theoretical construct
% LD: I think this is a problem because asymmetric representation is hypothesized to be an asymmetry in the way the two categories represented
Compared to \citeA{johansen2007paradoxical}, we do not consider asymmetric representation and base-rate neglect as necessary conditions.
From our point of view, they result from the experimental conditions, which was our focus here.

@@

We made available the two experiments written in javascript, the analysis code, the raw data, and all other supplementary materials.
Experiment 1 is shared via \href{https://osf.io/auwvt/?view_only=2dc8384074fa4bcf9f2e3937fdaee2b4}{OSF}, and \href{https://github.com/lenarddome/ply216-observational-ibre}{GitHub}.
Experiment 2 is shared via \href{https://osf.io/2tmc4/?view_only=489ebc888ef84a6b9b904072cfbe74df}{OSF} and \href{https://github.com/lenarddome/ply222-non-causal-ibre}{Github}.
The main repository that includes this manuscript and links to the materials for the two experiments can be found on \href{https://github.com/lenarddome/pu093-errorless-attentionless-ibre}{GitHub}.
